---
title: "Chapter 9 - Working with Dates"
output:
  html_document:
    toc: yes
    df_print: kable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Packages

```{r}
pacman::p_load(rio, lubridate, parsedate, zoo, tidyverse)
```

# Data

```{r}
linelist <- import("https://raw.githubusercontent.com/appliedepi/epirhandbook/main/data/linelist_cleaned.xlsx")
glimpse(linelist)
```

# Check what class the date columns are

```{r}
class(linelist$date_onset)      # datetime, not Date
class(linelist$time_admission)  # text like "09:36"
```

# Today's date

```{r}
Sys.Date()   # base R
Sys.time()
today()      # lubridate
now()
```

# Converting text to dates

## base R: as.Date() needs to be told the format

```{r}
as.Date("2020-10-11")                          # already year-month-day, no format needed
as.Date("11/10/2020", format = "%d/%m/%Y")     # %d = day, %m = month, %Y = 4-digit year
```

## lubridate: just say the order of the parts

```{r}
ymd("2020-10-11")
mdy("10/11/2020")
mdy("Oct 11 20")
dmy("11 10 2020")
dmy("11 October 2020")
```

## Converting the date columns in the linelist

```{r}
linelist <- linelist %>% 
  mutate(date_infection = as.Date(date_infection),
         date_onset = as.Date(date_onset),
         date_hospitalisation = as.Date(date_hospitalisation),
         date_outcome = as.Date(date_outcome))

class(linelist$date_onset)
```

# Excel dates

```{r}
as.Date(41772, origin = "1899-12-30")   # Excel stores dates as day counts from 30 Dec 1899
```

# Messy dates

```{r}
parsedate::parse_date(c("03 January 2018", "07/03/1982", "08/20/85"))   # guesses each format; check the guesses
```

# Dates with times

```{r}
ymd_h("2020-01-01 4PM")
dmy_hm("01 January 2020 16:20")
dmy_hms("01 January 2020, 16:20:40")   # the handbook uses mdy_hms() here and gets the wrong date
```

## Combining the date and time columns

```{r}
linelist <- linelist %>% 
  mutate(date_time_of_admission = ymd_hm(paste(date_hospitalisation, time_admission)))

head(linelist$date_time_of_admission)
```

## Time of day from the admission time

```{r}
linelist <- linelist %>% 
  mutate(hour_admit = hour(date_time_of_admission),
         time_period = case_when(hour_admit > 06 & hour_admit < 12 ~ "Morning",
                                 hour_admit >= 12 & hour_admit < 17 ~ "Afternoon",
                                 hour_admit >= 17 & hour_admit < 21 ~ "Evening",
                                 hour_admit >= 21 | hour_admit <= 6 ~ "Night"))

table(linelist$time_period, useNA = "always")
```

# Pulling pieces out of a date

```{r}
example_date <- ymd("2020-03-01")

year(example_date)
month(example_date)
day(example_date)
wday(example_date, label = TRUE)
```

# Date math

```{r}
example_date + days(3)
example_date + weeks(7) - days(2)
example_date - ymd("2020-02-20")                # difference between two dates
as.numeric(example_date - ymd("2020-02-20"))    # as a plain number
```

## Days from onset to hospitalisation

```{r}
linelist <- linelist %>% 
  mutate(days_onset_to_hosp = as.numeric(date_hospitalisation - date_onset))

median(linelist$days_onset_to_hosp, na.rm = TRUE)
```

# Changing how dates look

```{r}
format(Sys.Date(), "%d %B %Y")
format(Sys.Date(), "%A")
format(Sys.Date(), "%Y Week %W")
```

## Month-year that stays in the right order

```{r}
linelist <- linelist %>% 
  mutate(yearmonth = as.yearmon(date_onset))

table(linelist$yearmonth)
```

```{r}
table(format(linelist$date_onset, "%b %Y"))   # plain text sorts alphabetically instead
```

# Weekly case counts

```{r}
weekly_counts <- linelist %>% 
  drop_na(date_onset) %>% 
  mutate(week_onset = floor_date(date_onset, "week", week_start = 1)) %>%   # week_start = 1 means weeks start on Monday
  count(week_onset)

head(weekly_counts)
```

# Time zones

```{r}
time_now <- Sys.time()
time_now
with_tz(time_now, "Europe/London")    # same moment, London's clock
force_tz(time_now, "Europe/London")   # same clock time, relabelled as London
```

# Comparing each week to the one before

```{r}
weekly_counts <- weekly_counts %>% 
  mutate(cases_prev_wk = lag(n),
         case_diff = n - cases_prev_wk)

head(weekly_counts, 10)
```
