---
title: "Chapter 9 - Working with Dates"
output:
  html_document:
    toc: yes
    df_print: kable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Packages

```{r}
pacman::p_load(rio, lubridate, parsedate, zoo, tidyverse)
```

# Data

```{r}
linelist <- import("https://raw.githubusercontent.com/appliedepi/epirhandbook/main/data/linelist_cleaned.xlsx")
glimpse(linelist)
```

# Check what class the date columns are

```{r}
class(linelist$date_onset)      # datetime, not Date
class(linelist$time_admission)  # text like "09:36"
```

# Today's date

```{r}
Sys.Date()   # base R
Sys.time()
today()      # lubridate
now()
date()       # date and time as text
```

# Converting text to dates

## base R: as.Date() needs to be told the format

```{r}
as.Date("2020-10-11")                          # already year-month-day, no format needed
as.Date("11/10/2020", format = "%d/%m/%Y")     # %d = day, %m = month, %Y = 4-digit year
```

## lubridate: just say the order of the parts

```{r}
ymd("2020-10-11")
ymd("20201011")
mdy("10/11/2020")
mdy("Oct 11 20")
dmy("11 10 2020")
dmy("11 October 2020")
```

## Two-digit years

```{r}
two_digit_years <- c("15", "15", "16", "17")
paste0("20", two_digit_years)   # add the century first, then convert; the handbook uses str_glue() for this
```

## Converting the date columns in the linelist

```{r}
linelist <- linelist %>% 
  mutate(date_infection = as.Date(date_infection),
         date_onset = as.Date(date_onset),
         date_hospitalisation = as.Date(date_hospitalisation),
         date_outcome = as.Date(date_outcome))

class(linelist$date_onset)
```

## Building a date from separate year, month, and day columns

```{r}
make_date(year = 2014, month = 5, day = 13)   # works the same on columns, e.g. make_date(onset_year, onset_month, onset_day)
```

# Excel dates

```{r}
as.numeric(as.Date("1970-01-01"))       # R counts days from 1 Jan 1970
as.Date(41772, origin = "1899-12-30")   # Excel counts from 30 Dec 1899, so give it that origin instead of a format
```

# Messy dates

```{r}
parsedate::parse_date(c("03 January 2018", "07/03/1982", "08/20/85"))   # guesses each format; check the guesses
```

# Dates with times

```{r}
ymd_h("2020-01-01 16hrs")
ymd_h("2020-01-01 4PM")
dmy_hm("01 January 2020 16:20")
dmy_hms("01 January 2020, 16:20:40")       # the handbook uses mdy_hms() here and gets the wrong date
dmy_hms("01 January 2020, 16:20:40 PST")   # the time zone text is ignored
```

## Combining the date and time columns

```{r}
# The handbook first fills missing times with the median, but its median() line doesn't work on text, so missing times stay missing here
linelist <- linelist %>% 
  mutate(date_time_of_admission = ymd_hm(paste(date_hospitalisation, time_admission)))

head(linelist$date_time_of_admission)
```

## Times without dates

```{r}
time1 <- strptime("13:45", format = "%H:%M")   # today's date is assumed
time2 <- strptime("15:20", format = "%H:%M")

time2 - time1
as.numeric(time2 - time1)   # difference in hours, as a plain number
```

## Time of day from the admission time

```{r}
linelist <- linelist %>% 
  mutate(hour_admit = hour(strptime(time_admission, format = "%H:%M")),
         time_period = case_when(hour_admit > 06 & hour_admit < 12 ~ "Morning",
                                 hour_admit >= 12 & hour_admit < 17 ~ "Afternoon",
                                 hour_admit >= 17 & hour_admit < 21 ~ "Evening",
                                 hour_admit >= 21 | hour_admit <= 6 ~ "Night"))

table(linelist$time_period, useNA = "always")
```

# Pulling pieces out of a date

```{r}
example_date <- ymd("2020-03-01")
```

```{r}
month(example_date)
day(example_date)
wday(example_date)                 # day of the week as a number, Sunday = 1
wday(example_date, label = TRUE)
```

```{r}
example_datetime <- ymd_hm("2020-03-01 14:45")

hour(example_datetime)
minute(example_datetime)
second(example_datetime)
```

# Date math

```{r}
example_date + days(3)
example_date + weeks(7) - days(2)
```

## Difference between two dates

```{r}
output <- example_date - ymd("2020-02-20")
output
class(output)         # "difftime", not a plain number
as.numeric(output)    # now it's a number you can calculate with
```

## Days from onset to hospitalisation

```{r}
# The handbook's version subtracts date_of_onset, a column that doesn't exist
march_cases <- linelist %>% 
  filter(month(date_onset) == 3) %>%   # cases with onset in March
  mutate(days_onset_to_hosp = as.numeric(date_hospitalisation - date_onset))

median(march_cases$days_onset_to_hosp, na.rm = TRUE)   # na.rm = TRUE skips cases missing a date
```

# Changing how dates look

```{r}
format(Sys.Date(), "%d %B %Y")
format(Sys.Date(), "%A")
format(Sys.Date(), "%Y Week %W")   # format() gives back text, so use it for display only
```

## Month-year that stays in the right order

```{r}
linelist <- linelist %>% 
  mutate(yearmonth = as.yearmon(date_onset))

table(linelist$yearmonth)   # the handbook writes $yearmon, which only works by accident
```

```{r}
table(format(linelist$date_onset, "%b %Y"))   # plain text sorts alphabetically instead
```

# Weekly case counts

```{r}
weekly_counts <- linelist %>% 
  drop_na(date_onset) %>%                                                   # drop cases with no onset date
  mutate(week_onset = floor_date(date_onset, "week", week_start = 1)) %>%   # week_start = 1 means Monday; the default is Sunday, not Monday as the handbook says
  count(week_onset, name = "cases_wk") %>% 
  complete(week_onset = seq.Date(min(week_onset), max(week_onset), by = "week"),
           fill = list(cases_wk = 0))                                        # add any weeks with zero cases

head(weekly_counts)
```

# Time zones

```{r}
time_now <- Sys.time()
time_london_real <- with_tz(time_now, "Europe/London")     # same moment, London's clock
time_london_local <- force_tz(time_now, "Europe/London")   # same clock time, relabelled as London

time_now
time_london_real
time_london_local
time_london_real - time_london_local   # the gap between your time zone and London
```

# Comparing each week to the one before

```{r}
weekly_counts <- weekly_counts %>% 
  mutate(cases_prev_wk = lag(cases_wk, n = 1),   # the previous row's count
         case_diff = cases_wk - cases_prev_wk)

head(weekly_counts, 10)
```

# Resources

```{r}
# lubridate: https://lubridate.tidyverse.org/
# lubridate cheatsheet: https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf
# R for Data Science, dates and times: https://r4ds.had.co.nz/dates-and-times.html
# Online tutorial: https://www.statmethods.net/input/dates.html
# Date formats: https://www.r-bloggers.com/2013/08/date-formats-in-r/
```
