---
title: "Chapter 9 - Working with Dates (handbook verbatim)"
output:
  html_document:
    toc: yes
    df_print: kable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r notes}
# HANDBOOK VERBATIM. The code below is copied from the handbook's source file (chapters/dates.qmd)
# with the handbook's own comments and chunk options. Chunks marked eval=FALSE are displayed but not
# run, exactly as on the handbook web page. The only departures, all needed to knit it in this project:
#   - Files are loaded with here("Data", ...) instead of a bare file name in the working directory.
#   - The handbook runs on linelist_cleaned.rds (date columns already class Date) while displaying an
#     .xlsx import it never executes. This file imports the .rds. If you only have the .xlsx, add
#     mutate(across(where(is.POSIXct), as.Date)) after import(), or seq.Date() in the weekly counts fails.
#   - Two chunks the handbook hides are shown: the .rds import, and the construction of `counts` from
#     example/district_weekly_count_data.xlsx, which the lag section needs.
#   - The handbook's DT::datatable() display chunks are replaced with head(). The header image is omitted.
#
# Things in the handbook's own text or code that are wrong, left untouched below:
#   - floor_date()'s default week_start is 7 (Sunday), not 1 (Monday) as the text says.
#   - mdy_hms("01 January 2020, 16:20:40") is a day-month-year string, so it silently returns 2020-01-20.
#   - median(time_admission) on character text returns NA, so the "fill missing times with the median"
#     step does nothing (eval=FALSE chunk).
#   - The delay example subtracts date_of_onset, a column that does not exist (eval=FALSE chunk).
#   - The strptime list gives %h, %m, %s for hours, minutes, seconds; the working codes are %H, %M, %S.
#   - table(test_zoo$yearmon) works only because the data are a data.frame and $ partial-matches the
#     column "yearmonth"; on a tibble it returns an empty table.
#   - Once tidyverse is attached, an unqualified parse_date() is readr's, not parsedate's. The vector
#     example uses the parsedate:: prefix; the column example (eval=FALSE) does not.
```

# Preparation

## Load packages

```{r dates_packages, warning=F, message=F}
# Checks if package is installed, installs if necessary, and loads package for current session

pacman::p_load(
  lubridate,  # general package for handling and converting dates  
  parsedate,  # has function to "guess" messy dates
  aweek,      # another option for converting dates to weeks, and weeks to dates
  zoo,        # additional date/time functions
  here,       # file management
  rio,        # data import/export
  tidyverse)  # data management and visualization  
```

## Import data

```{r import-rds}
# (hidden chunk in the handbook: this is the file the page actually runs on)
linelist <- rio::import(here::here("Data", "linelist_cleaned.rds"))
```

```{r import-xlsx, eval=F}
linelist <- import("linelist_cleaned.xlsx")
```

# Current date

```{r current-date}
# get the system date - this is a DATE class
Sys.Date()

# get the system time - this is a DATETIME class
Sys.time()
```

# Convert to Date

## base R

```{r as-date, eval=F}
# Convert to class date
linelist <- linelist %>% 
  mutate(date_onset = as.Date(date_of_onset, format = "%d/%m/%Y"))
```

## lubridate

```{r load-lubridate}
# install/load lubridate 
pacman::p_load(lubridate)
```

```{r ymd}
# read date in year-month-day format
ymd("2020-10-11")
ymd("20201011")
```

```{r mdy}
# read date in month-day-year format
mdy("10/11/2020")
mdy("Oct 11 20")
```

```{r dmy}
# read date in day-month-year format
dmy("11 10 2020")
dmy("11 October 2020")
```

```{r dmy-column, eval=F}
linelist <- linelist %>%
  mutate(date_onset = lubridate::dmy(date_onset))
```

```{r check-class, eval=F}
# Check the class of the column
class(linelist$date_onset)  
```

```{r two-digit-years}
two_digit_years <- c("15", "15", "16", "17")
str_glue("20{two_digit_years}")
```

## Combine columns

```{r make-date, eval=F}
linelist <- linelist %>% 
  mutate(onset_date = make_date(year = onset_year, month = onset_month, day = onset_day))
```

# Excel dates

```{r excel-dates, eval = FALSE}
# An example of providing the Excel 'origin date' when converting Excel number dates
data_cleaned <- data %>% 
  mutate(date_onset = as.numeric(date_onset)) %>%   # ensure class is numeric
  mutate(date_onset = as.Date(date_onset, origin = "1899-12-30")) # convert to date using Excel origin
```

# Messy dates

```{r parse-date}
parsedate::parse_date(c("03 January 2018",
                        "07/03/1982",
                        "08/20/85"))
```

```{r parse-date-column, eval = FALSE}
# An example using parse_date() on the column date_onset
linelist <- linelist %>%      
  mutate(date_onset = parse_date(date_onset))
```

# Working with date-time class

## Convert dates with times

```{r ymd-h}
ymd_h("2020-01-01 16hrs")
ymd_h("2020-01-01 4PM")
```

```{r dmy-hm}
dmy_hm("01 January 2020 16:20")
```

```{r mdy-hms}
mdy_hms("01 January 2020, 16:20:40")
```

```{r mdy-hms-tz}
mdy_hms("01 January 2020, 16:20:40 PST")

```

```{r combine-date-time, eval = FALSE}
# packages
pacman::p_load(tidyverse, lubridate, stringr)

# time_admission is a column in hours:minutes
linelist <- linelist %>%
  
  # when time of admission is not given, assign the median admission time
  mutate(
    time_admission_clean = ifelse(
      is.na(time_admission),         # if time is missing
      median(time_admission),        # assign the median
      time_admission                 # if not missing keep as is
  ) %>%
  
    # use str_glue() to combine date and time columns to create one character column
    # and then use ymd_hm() to convert it to datetime
  mutate(
    date_time_of_admission = str_glue("{date_hospitalisation} {time_admission_clean}") %>% 
      ymd_hm()
  )

```

## Convert times alone

```{r times-alone}
# raw character times
time1 <- "13:45" 
time2 <- "15:20"

# Times converted to a datetime class
time1_clean <- strptime(time1, format = "%H:%M")
time2_clean <- strptime(time2, format = "%H:%M")

# Difference is of class "difftime" by default, here converted to numeric hours 
as.numeric(time2_clean - time1_clean)   # difference in hours

```

## Extract time

```{r extract-time}
linelist <- linelist %>%
  mutate(hour_admit = hour(strptime(time_admission, format = "%H:%M"))) %>%
  mutate(time_period = case_when(
    hour_admit > 06 & hour_admit < 12 ~ "Morning",
    hour_admit >= 12 & hour_admit < 17 ~ "Afternoon",
    hour_admit >= 17 & hour_admit < 21 ~ "Evening",
    hour_admit >=21 | hour_admit <= 6 ~ "Night"))
```

# Working with dates

```{r example-date}
# create object of class Date
example_date <- ymd("2020-03-01")
```

## Extract date components

```{r date-components}
month(example_date)  # month number
day(example_date)    # day (number) of the month
wday(example_date)   # day number of the week (1-7)
```

```{r time-components, eval=F}
example_datetime <- ymd_hm("2020-03-01 14:45")

hour(example_datetime)     # extract hour
minute(example_datetime)   # extract minute
second(example_datetime)   # extract second
```

## Date math

```{r date-math}
# add 3 days to this date
example_date + days(3)
  
# add 7 weeks and subtract two days from this date
example_date + weeks(7) - days(2)
```

## Date intervals

```{r date-intervals}
# find the interval between this date and Feb 20 2020 
output <- example_date - ymd("2020-02-20")
output    # print
class(output)
```

```{r delay, eval = F}
pacman::p_load(lubridate, tidyverse)   # load packages

linelist <- linelist %>%
  
  # convert date of onset from character to date objects by specifying dmy format
  mutate(date_onset = dmy(date_onset),
         date_hospitalisation = dmy(date_hospitalisation)) %>%
  
  # filter out all cases without onset in march
  filter(month(date_onset) == 3) %>%
    
  # find the difference in days between onset and hospitalisation
  mutate(days_onset_to_hosp = date_hospitalisation - date_of_onset)
```

```{r delay-median, eval = FALSE}
# calculate the median number of days to hospitalisation for all cases where data are available
median(linelist_delay$days_onset_to_hosp, na.rm = T)
```

# Date display

## format()

```{r format}
# today's date, with formatting
format(Sys.Date(), format = "%d %B %Y")

# easy way to get full date and time (default formatting)
date()

# formatted combined date, time, and time zone using str_glue() function
str_glue("{format(Sys.Date(), format = '%A, %B %d %Y, %z  %Z, ')}{format(Sys.time(), format = '%H:%M:%S')}")

# Using format to display weeks
format(Sys.Date(), "%Y Week %W")
```

## Month-Year

```{r yearmon}
# create new column 
test_zoo <- linelist %>% 
     mutate(yearmonth = zoo::as.yearmon(date_onset))

# print table
table(test_zoo$yearmon)
```

```{r format-yearmon}
# create new column
test_format <- linelist %>% 
     mutate(yearmonth = format(date_onset, "%b %Y"))

# print table
table(test_format$yearmon)
```

# Epidemiological weeks

## Weekly counts

```{r weekly-counts}
# Make aggregated dataset of weekly case counts
weekly_counts <- linelist %>% 
  drop_na(date_onset) %>%             # remove cases missing onset date
  mutate(weekly_cases = floor_date(   # make new column, week of onset
    date_onset,
    unit = "week")) %>%            
  count(weekly_cases) %>%           # group data by week and count rows per group (creates column 'n')
  tidyr::complete(                  # ensure all weeks are present, even those with no cases reported
    weekly_cases = seq.Date(          # re-define the "weekly_cases" column as a complete sequence,
      from = min(weekly_cases),       # from the minimum date
      to = max(weekly_cases),         # to the maxiumum date
      by = "week"),                   # by weeks
    fill = list(n = 0))             # fill-in NAs in the n counts column with 0
```

```{r weekly-counts-view}
# (the handbook shows this with DT::datatable())
head(weekly_counts, 20)
```

# Converting dates/time zones

```{r time-zones}
# assign the current time to a column
time_now <- Sys.time()
time_now

# use with_tz() to assign a new timezone to the column, while CHANGING the clock time
time_london_real <- with_tz(time_now, "Europe/London")

# use force_tz() to assign a new timezone to the column, while KEEPING the clock time
time_london_local <- force_tz(time_now, "Europe/London")


# note that as long as the computer that was used to run this code is NOT set to London time,
# there will be a difference in the times 
# (the number of hours difference from the computers time zone to london)
time_london_real - time_london_local

```

# Lagging and leading calculations

```{r counts-data}
# (hidden chunk in the handbook: builds the weekly counts used below)
counts <- import(here("Data", "example", "district_weekly_count_data.xlsx")) %>% 
  filter(District == "Nibari") %>% 
  mutate(Date = as.Date(Date),
         week_start = lubridate::floor_date(Date, "week")) %>%
  group_by(week_start) %>% 
  summarize(cases_wk = sum(Cases, na.rm=T)) %>% 
  complete(week_start = seq.Date(min(week_start), max(week_start), by = "week"), fill = list(cases_wk = 0))
```

```{r counts-view}
# (the handbook shows this with DT::datatable())
head(counts)
```

```{r lag}
counts <- counts %>% 
  mutate(cases_prev_wk = lag(cases_wk, n = 1))
```

```{r lag-view}
head(counts)
```

```{r lag-diff}
counts <- counts %>% 
  mutate(cases_prev_wk = lag(cases_wk, n = 1),
         case_diff = cases_wk - cases_prev_wk)
```

```{r lag-diff-view}
head(counts)
```

# Resources

```{r resources}
# lubridate tidyverse page: https://lubridate.tidyverse.org/
# lubridate RStudio cheatsheet: https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf
# R for Data Science, dates and times: https://r4ds.had.co.nz/dates-and-times.html
# Online tutorial: https://www.statmethods.net/input/dates.html
# Date formats: https://www.r-bloggers.com/2013/08/date-formats-in-r/
```
