---
title: "Chapter 9 - Working with Dates"
output:
  html_document:
    toc: yes
    df_print: kable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Connect to libraries

```{r packages}
pacman::p_load(rio, here, lubridate, parsedate, zoo, tidyverse, magrittr)
```

# Connect to data

```{r load-data}
linelist <- here("Data", "linelist_cleaned.xlsx") %>% import()
glimpse(linelist)
```

# Review the date columns

```{r check-classes}
class(linelist$date_onset) # Imported from Excel as POSIXct (datetime), not Date
class(linelist$time_admission) # Character text in HH:MM format
```

# Current date

## Get the system date and time with base R

```{r sys-date}
Sys.Date() # Date class
Sys.time() # Datetime class
```

## Get the system date and time with lubridate

```{r lubridate-today}
today()
now()
date()
```

# Converting to Date class

## Converting with base R as.Date()

```{r as-date-column}
linelist %>% 
  mutate(date_onset = as.Date(date_onset)) %>% # Datetime to Date; no format needed
  select(date_onset) %>% 
  glimpse()
```

```{r as-date-format-column}
linelist %>% 
  mutate(date_onset_raw = format(date_onset, "%d/%m/%Y"), # Simulate a raw import: the date as DD/MM/YYYY text
         date_onset_new = as.Date(date_onset_raw, format = "%d/%m/%Y")) %>% # format describes the text as it is BEFORE conversion
  select(date_onset, date_onset_raw, date_onset_new) %>% 
  glimpse()
```

## Converting with lubridate::ymd(), mdy(), and dmy()

```{r lubridate-parsers}
ymd("2020-10-11") # year-month-day
ymd("20201011")
mdy("10/11/2020") # month-day-year
mdy("Oct 11 20")
dmy("11 10 2020") # day-month-year
dmy("11 October 2020")
```

```{r lubridate-parsers-column}
linelist %>% 
  mutate(date_onset_raw = format(date_onset, "%d/%m/%Y"),
         date_onset_new = dmy(date_onset_raw)) %>% # Only the ORDER of the parts matters, not the separators
  select(date_onset, date_onset_raw, date_onset_new) %>% 
  glimpse()
```

## Handling two-digit years

```{r two-digit-years}
two_digit_years <- c("15", "15", "16", "17")
str_glue("20{two_digit_years}") # Add the century first, then convert to Date
```

## Combining separate columns with make_date()

```{r make-date}
linelist %>% 
  mutate(onset_year = year(date_onset), # Split the date into parts to simulate data that arrive as separate columns
         onset_month = month(date_onset),
         onset_day = day(date_onset),
         date_onset_new = make_date(year = onset_year, month = onset_month, day = onset_day)) %>% # Recombine into one Date column
  select(date_onset, onset_year, onset_month, onset_day, date_onset_new) %>% 
  glimpse()
```

## Applying date class conversions

```{r apply-date-class}
linelist <- here("Data", "linelist_cleaned.xlsx") %>% 
  import() %>% 
  mutate(across(where(is.POSIXct), as.Date)) # Convert every datetime column to Date

linelist %>% 
  select(contains("date")) %>% 
  glimpse()
```

# Excel dates

```{r excel-dates}
as.numeric(as.Date("1970-01-01")) # R counts days from 1 Jan 1970, so this is day 0

linelist %>% 
  mutate(date_onset_excel = as.numeric(date_onset - as.Date("1899-12-30")), # Simulate the serial numbers Excel stores (Windows origin is 30 Dec 1899)
         date_onset_new = as.Date(date_onset_excel, origin = "1899-12-30")) %>% # Supply the origin instead of a format
  select(date_onset, date_onset_excel, date_onset_new) %>% 
  head()
```

# Messy dates

## Guessing formats with parsedate::parse_date()

```{r parsedate-vector}
parsedate::parse_date(c("03 January 2018", "07/03/1982", "08/20/85")) # Check the guesses: "07/03/1982" became July 3rd
```

```{r parsedate-column}
# readr (loaded by tidyverse) also has a parse_date(), so use the parsedate:: prefix
linelist %>% 
  drop_na(date_onset) %>% 
  mutate(date_onset_messy = if_else(row_number() %% 2 == 0, # Simulate a column with two different formats
                                    format(date_onset, "%d %B %Y"),
                                    format(date_onset, "%Y/%m/%d")),
         date_onset_new = as.Date(parsedate::parse_date(date_onset_messy))) %>% # Guess each row's format; as.Date() drops the time part
  select(date_onset, date_onset_messy, date_onset_new) %>% 
  head()
```

# Working with date-time class

## Converting datetime text with lubridate

```{r datetime-parsers}
ymd_h("2020-01-01 16hrs") # Date plus hours
ymd_h("2020-01-01 4PM")
dmy_hm("01 January 2020 16:20") # Date plus hours and minutes
dmy_hms("01 January 2020, 16:20:40") # The handbook uses mdy_hms() here, which silently returns 2020-01-20; match the parser to the order of the parts
dmy_hms("01 January 2020, 16:20:40 PST") # Time zone text is ignored
```

## Combining a date column and a time column

```{r median-time}
median_time <- linelist$time_admission %>% 
  strptime(format = "%H:%M") %>% # HH:MM text to datetime; impossible values like "12:60" become NA
  as.POSIXct() %>% 
  median(na.rm = TRUE) %>% # The handbook's median(time_admission) on the raw text returns NA
  format("%H:%M") # Back to HH:MM text

median_time
```

```{r combine-date-time}
linelist %<>%
  mutate(time_admission_clean = replace_na(time_admission, median_time), # Fill missing times with the median
         date_time_of_admission = ymd_hm(str_glue("{date_hospitalisation} {time_admission_clean}"))) # Paste date and time, then convert; impossible times become NA

linelist %>% 
  select(date_hospitalisation, time_admission, time_admission_clean, date_time_of_admission) %>% 
  glimpse()
```

## Converting times alone with strptime()

```{r times-alone}
time1 <- "13:45"
time2 <- "15:20"

time1_clean <- strptime(time1, format = "%H:%M") # Today's date is assumed
time2_clean <- strptime(time2, format = "%H:%M")

as.numeric(time2_clean - time1_clean) # Difference in hours
```

## Extracting time components and classifying by time of day

```{r hour-and-period}
linelist %<>%
  mutate(hour_admit = hour(strptime(time_admission, format = "%H:%M")), # Pull out the hour (0-23)
         time_period = case_when(hour_admit > 06 & hour_admit < 12 ~ "Morning",
                                 hour_admit >= 12 & hour_admit < 17 ~ "Afternoon",
                                 hour_admit >= 17 & hour_admit < 21 ~ "Evening",
                                 hour_admit >= 21 | hour_admit <= 6 ~ "Night"))

linelist %$% 
  table(time_period, useNA = "always")
```

# Working with dates

## Extracting date components

```{r date-components}
example_date <- ymd("2020-03-01")

month(example_date) # Month number
day(example_date) # Day of the month
wday(example_date) # Day of the week (1 = Sunday)
```

```{r time-components}
example_datetime <- ymd_hm("2020-03-01 14:45")

hour(example_datetime)
minute(example_datetime)
second(example_datetime)
```

## Date math

```{r date-math}
example_date + days(3) # Add 3 days
example_date + weeks(7) - days(2) # Add 7 weeks and subtract 2 days
```

## Date intervals

```{r date-intervals}
output <- example_date - ymd("2020-02-20")
output
class(output) # difftime, not a number
as.numeric(output) # Convert before using in other calculations
```

## Calculating delays between two date columns

```{r delay-march}
linelist_delay <- linelist %>% 
  filter(month(date_onset) == 3) %>% # Keep only cases with onset in March
  mutate(days_onset_to_hosp = as.numeric(date_hospitalisation - date_onset)) # Rows missing either date get NA

linelist_delay %>% 
  select(date_onset, date_hospitalisation, days_onset_to_hosp) %>% 
  glimpse()

median(linelist_delay$days_onset_to_hosp, na.rm = TRUE)
```

# Date display

## Changing the display with format()

```{r format-display}
format(Sys.Date(), format = "%d %B %Y") # Returns character, so use for display only
date()
str_glue("{format(Sys.Date(), format = '%A, %B %d %Y, %z  %Z, ')}{format(Sys.time(), format = '%H:%M:%S')}") # Single quotes inside str_glue()'s double quotes
format(Sys.Date(), "%Y Week %W")
```

## Month-year with zoo::as.yearmon()

```{r yearmon}
test_zoo <- linelist %>% 
  mutate(yearmonth = zoo::as.yearmon(date_onset)) # Class yearmon keeps chronological order

table(test_zoo$yearmonth) # Full column name; the handbook's $yearmon only works by partial matching
```

```{r format-yearmon}
test_format <- linelist %>% 
  mutate(yearmonth = format(date_onset, "%b %Y")) # Class character sorts alphabetically instead

table(test_format$yearmonth)
```

# Epidemiological weeks

## Weekly case counts with floor_date()

```{r weekly-counts}
weekly_counts <- linelist %>% 
  drop_na(date_onset) %>% # Remove cases missing onset date
  mutate(weekly_cases = floor_date(date_onset, unit = "week", week_start = 1)) %>% # Week of onset; lubridate's default week_start is 7 (Sunday), not Monday as the handbook says
  count(weekly_cases) %>% # Rows per week, as column n
  tidyr::complete(weekly_cases = seq.Date(from = min(weekly_cases), # Add any weeks with no cases
                                          to = max(weekly_cases),
                                          by = "week"),
                  fill = list(n = 0)) # Give them 0 instead of NA

head(weekly_counts)
```

# Converting dates/time zones

```{r time-zones}
time_now <- Sys.time()
time_now

time_london_real <- with_tz(time_now, "Europe/London") # Same instant, clock time changes
time_london_local <- force_tz(time_now, "Europe/London") # Same clock time, instant changes

time_london_real - time_london_local # Offset between the local time zone and London
```

# Lagging and leading calculations

```{r lag-lead}
weekly_counts %<>%
  mutate(cases_prev_wk = lag(n, n = 1), # Previous row's count; the first row gets NA
         case_diff = n - cases_prev_wk) # Change since the previous week (rows must be sorted by date first)

head(weekly_counts, 10)
```
