---
title: "Chapter 9 - Working with Dates"
output: html_notebook
---

# Connect to libraries

```{r}
pacman::p_load(rio, here, lubridate, parsedate, aweek, zoo, tidyverse, magrittr)
```

# Connect to data

```{r}
linelist <- here("Data", "linelist_cleaned.xlsx") %>% import()
glimpse(linelist)
```

# Review the date columns

## Check the class of the date and time columns

```{r}
class(linelist$date_onset) # Excel stores dates as datetimes, so rio::import() brings them in as POSIXct (datetime), not Date
class(linelist$time_admission) # Admission time is character text in HH:MM format

linelist %>% 
  select(where(is.POSIXct)) %>% # all the datetime columns
  names()
```

# Current date

## Get the system date and time with base R

```{r}
Sys.Date() # Date class
Sys.time() # Datetime class (POSIXct), in the computer's local time zone
```

## Get the system date and time with lubridate

```{r}
today() # Date class
now() # Datetime class
date() # Character string with the weekday and month names
```

# Converting to Date class

## Converting a datetime column with as.Date()

```{r}
linelist %>% 
  mutate(date_onset = as.Date(date_onset)) %>% # POSIXct to Date; no format needed because R already knows these are datetimes
  select(date_onset) %>% 
  glimpse()
```

## Converting character dates with as.Date() and a format string

```{r}
as.Date("1968-04-24") # Already in R's standard YYYY-MM-DD format, so no format argument is needed
as.Date("24/04/1968", format = "%d/%m/%Y") # format describes how the raw text looks BEFORE conversion, not how we want it displayed
as.Date("24/04/1968") # Without the format R silently reads this as year 24, so always check the result
```

```{r}
linelist %>% 
  mutate(date_onset_raw = format(date_onset, "%d/%m/%Y"), # Simulate a raw import: the date as DD/MM/YYYY text
         date_onset_new = as.Date(date_onset_raw, format = "%d/%m/%Y")) %>% # Tell R which piece is the day, month, and year
  select(date_onset, date_onset_raw, date_onset_new) %>% 
  glimpse()
```

## strptime abbreviations used by format arguments

```{r}
format(Sys.Date(), "%d") # Day of the month
format(Sys.Date(), "%j") # Day of the year
format(Sys.Date(), "%a %A") # Abbreviated and full weekday
format(Sys.Date(), "%u %w") # Weekday number: Monday = 1 (%u) or Sunday = 0 (%w)
format(Sys.Date(), "%W %U") # Week of the year: Monday start (%W) or Sunday start (%U)
format(Sys.Date(), "%m %b %B") # Month number, abbreviated month, and full month
format(Sys.Date(), "%y %Y") # 2-digit and 4-digit year
format(Sys.time(), "%H:%M:%S %z %Z") # Hours, minutes, seconds, offset from GMT, and time zone (the handbook's table lists these in lowercase, but they are uppercase)
```

## Converting with lubridate::ymd(), mdy(), and dmy()

```{r}
ymd("2020-10-11") # year-month-day
ymd("20201011")
mdy("10/11/2020") # month-day-year
mdy("Oct 11 20")
dmy("11 10 2020") # day-month-year
dmy("11 October 2020")
```

```{r}
linelist %>% 
  mutate(date_onset_raw = format(date_onset, "%d/%m/%Y"),
         date_onset_new = dmy(date_onset_raw)) %>% # lubridate only needs the ORDER of the parts, not the exact separators
  select(date_onset, date_onset_raw, date_onset_new) %>% 
  glimpse()
```

## Handling two-digit years

```{r}
dmy("11 10 20") # lubridate has to guess the century, so check the result

two_digit_years <- c("15", "15", "16", "17")
four_digit_years <- str_glue("20{two_digit_years}") # Paste a century prefix onto each value first
four_digit_years
```

## Combining separate columns with make_date()

```{r}
linelist %>% 
  mutate(onset_year = year(date_onset), # Split the date into parts to simulate data that arrive as separate columns
         onset_month = month(date_onset),
         onset_day = day(date_onset),
         date_onset_new = make_date(year = onset_year, month = onset_month, day = onset_day)) %>% # Recombine the parts into one Date column
  select(date_onset, onset_year, onset_month, onset_day, date_onset_new) %>% 
  glimpse()
```

## Applying date class conversions

```{r}
linelist <- here("Data", "linelist_cleaned.xlsx") %>% 
  import() %>% 
  mutate(across(where(is.POSIXct), as.Date)) # Convert every datetime column to Date in one step

linelist %>% 
  select(contains("date")) %>% 
  glimpse()
```

# Excel dates

## Origin dates

```{r}
as.numeric(as.Date("1970-01-01")) # R counts days from 1 Jan 1970, so its origin is day 0
as.Date(41772, origin = "1899-12-30") # Excel on Windows counts from 30 Dec 1899; 41772 is what Excel stored for 2014-05-13
as_date(41772, origin = "1899-12-30") # lubridate equivalent
```

## Converting a column of Excel serial numbers

```{r}
linelist %>% 
  mutate(date_onset_excel = as.numeric(date_onset - as.Date("1899-12-30")), # Simulate the serial numbers Excel stores behind the scenes
         date_onset_new = as.Date(date_onset_excel, origin = "1899-12-30")) %>% # Convert back by supplying the Excel origin instead of a format
  select(date_onset, date_onset_excel, date_onset_new) %>% 
  glimpse()
```

# Messy dates

## Guessing formats with parsedate::parse_date()

```{r}
parsedate::parse_date(c("03 January 2018", "07/03/1982", "08/20/85")) # Returns POSIXct in UTC; "07/03/1982" was read as July 3rd, not March 7th, so check the guesses
```

```{r}
# readr (loaded by tidyverse) also has a parse_date() function, so always write parsedate::parse_date() to get the right one
linelist %>% 
  drop_na(date_onset) %>% 
  mutate(date_onset_messy = if_else(row_number() %% 2 == 0, # Simulate a column where every other row uses a different format
                                    format(date_onset, "%d %B %Y"),
                                    format(date_onset, "%Y/%m/%d")),
         date_onset_new = as.Date(parsedate::parse_date(date_onset_messy))) %>% # parse_date() guesses each row's format; as.Date() drops the time component
  select(date_onset, date_onset_messy, date_onset_new) %>% 
  head()
```

# Working with date-time class

## Converting datetime text with lubridate

```{r}
ymd_h("2020-01-01 16hrs") # Date plus hours
ymd_h("2020-01-01 4PM")
dmy_hm("01 January 2020 16:20") # Date plus hours and minutes
dmy_hms("01 January 2020, 16:20:40") # Date plus hours, minutes, and seconds
dmy_hms("01 January 2020, 16:20:40 PST") # The time zone text is ignored; the result is still UTC
mdy_hms("01 January 2020, 16:20:40") # The handbook uses mdy_hms() here, but the parts are in day-month-year order, so the result is silently wrong (2020-01-20)
```

## Combining a date column and a time column

```{r}
median_time <- linelist$time_admission %>% 
  strptime(format = "%H:%M") %>% # Convert the HH:MM text to datetimes (today's date is assumed); impossible values like "12:60" become NA
  as.POSIXct() %>% 
  median(na.rm = TRUE) %>% # median() cannot work on the raw text, so the handbook's median(time_admission) just returns NA
  format("%H:%M") # Back to HH:MM text

median_time
```

```{r}
linelist %<>%
  mutate(time_admission_clean = replace_na(time_admission, median_time), # Fill missing admission times with the median time
         date_time_of_admission = ymd_hm(str_glue("{date_hospitalisation} {time_admission_clean}"))) # Paste date and time together, then convert; rows with impossible times such as "12:60" fail to parse and become NA

linelist %>% 
  select(date_hospitalisation, time_admission, time_admission_clean, date_time_of_admission) %>% 
  glimpse()
```

## Converting times alone with strptime()

```{r}
time1 <- "13:45"
time2 <- "15:20"

time1_clean <- strptime(time1, format = "%H:%M") # Without a date, today's date is assumed
time2_clean <- strptime(time2, format = "%H:%M")

time2_clean - time1_clean # Class difftime; R picks the units
as.numeric(time2_clean - time1_clean) # Difference in hours, as a plain number
difftime(time2_clean, time1_clean, units = "mins") # Set the units explicitly
```

## Extracting time components and classifying by time of day

```{r}
linelist %<>%
  mutate(hour_admit = hour(strptime(time_admission, format = "%H:%M")), # Convert to datetime, then pull out the hour (0-23)
         time_period = case_when(hour_admit > 06 & hour_admit < 12 ~ "Morning",
                                 hour_admit >= 12 & hour_admit < 17 ~ "Afternoon",
                                 hour_admit >= 17 & hour_admit < 21 ~ "Evening",
                                 hour_admit >= 21 | hour_admit <= 6 ~ "Night"))

linelist %$% 
  table(time_period, useNA = "always")
```

# Working with dates

## Extracting date components

```{r}
example_date <- ymd("2020-03-01")

year(example_date)
month(example_date) # Month number
month(example_date, label = TRUE) # Month name
day(example_date) # Day of the month
wday(example_date) # Day of the week as a number; lubridate counts from Sunday = 1 by default
wday(example_date, label = TRUE)
```

```{r}
example_datetime <- ymd_hm("2020-03-01 14:45")

hour(example_datetime)
minute(example_datetime)
second(example_datetime)
```

## Date math

```{r}
example_date + days(3) # Add 3 days
example_date + weeks(7) - days(2) # Add 7 weeks and subtract 2 days
example_date + months(1) # Add 1 month
```

## Date intervals

```{r}
output <- example_date - ymd("2020-02-20") # Subtract two Dates
output
class(output) # difftime, not a plain number
as.numeric(output) # Convert to numeric before using the result in other calculations
```

## Calculating delays between two date columns

```{r}
linelist_delay <- linelist %>% 
  filter(month(date_onset) == 3) %>% # Keep only cases with onset in March
  mutate(days_onset_to_hosp = as.numeric(date_hospitalisation - date_onset)) # Difference in days; rows missing either date get NA

linelist_delay %>% 
  select(date_onset, date_hospitalisation, days_onset_to_hosp) %>% 
  glimpse()

median(linelist_delay$days_onset_to_hosp, na.rm = TRUE) # na.rm = TRUE ignores the rows where a date was missing
```

## Applying date calculations

```{r}
linelist %<>%
  mutate(days_onset_to_hosp = as.numeric(date_hospitalisation - date_onset), # Same as days_onset_hosp from the chapter 8 cleaning, recalculated here for practice
         days_hosp_to_outcome = as.numeric(date_outcome - date_hospitalisation)) # Negative values here mean the outcome date is before the hospitalisation date, which is worth checking in the raw data

linelist %>% 
  select(days_onset_hosp, days_onset_to_hosp, days_hosp_to_outcome) %>% 
  summary()
```

# Date display

## Changing the display with format()

```{r}
format(Sys.Date(), format = "%d %B %Y") # Day, full month name, 4-digit year
date() # Full date and time with default formatting
str_glue("{format(Sys.Date(), format = '%A, %B %d %Y, %z  %Z, ')}{format(Sys.time(), format = '%H:%M:%S')}") # Use single quotes inside str_glue()'s double quotes
format(Sys.Date(), "%Y Week %W") # Year and week number

class(format(Sys.Date(), format = "%d %B %Y")) # format() returns character, so use it for display only, not for calculations
```

```{r}
linelist %>% 
  mutate(date_onset_display = format(date_onset, "%A %d %B %Y")) %>% 
  select(date_onset, date_onset_display) %>% 
  head()
```

## Month-year with zoo::as.yearmon()

```{r}
test_zoo <- linelist %>% 
  mutate(yearmonth = zoo::as.yearmon(date_onset)) # Class yearmon keeps chronological order

table(test_zoo$yearmonth)
```

```{r}
test_format <- linelist %>% 
  mutate(yearmonth = format(date_onset, "%b %Y")) # Class character sorts alphabetically instead

table(test_format$yearmonth)
```

# Epidemiological weeks

## Rounding dates down to the week with floor_date()

```{r}
linelist %>% 
  mutate(week_onset = floor_date(date_onset, unit = "week", week_start = 1)) %>% # Start of the week (Monday = 1); lubridate's default week_start is 7 (Sunday), not Monday as the handbook text says
  select(date_onset, week_onset) %>% 
  head()
```

## Weekly case counts

```{r}
weekly_counts <- linelist %>% 
  drop_na(date_onset) %>% # Remove cases missing onset date
  mutate(weekly_cases = floor_date(date_onset, unit = "week", week_start = 1)) %>% # Week of onset
  count(weekly_cases) %>% # Rows per week, as column n
  tidyr::complete(weekly_cases = seq.Date(from = min(weekly_cases), # Add any weeks with no cases
                                          to = max(weekly_cases),
                                          by = "week"),
                  fill = list(n = 0)) # Fill their counts with 0 instead of NA

head(weekly_counts)
```

## Epiweek alternatives

```{r}
linelist %>% 
  drop_na(date_onset) %>% 
  mutate(week = week(date_onset), # Week of the year counted in 7-day blocks from 1 January
         epiweek = epiweek(date_onset), # US CDC epi week (Sunday start)
         isoweek = isoweek(date_onset), # ISO 8601 week (Monday start)
         aweek = as.character(date2week(date_onset, week_start = "Monday", floor_day = TRUE))) %>% # "YYYY-Www" text from the aweek package
  select(date_onset, week, epiweek, isoweek, aweek) %>% 
  head()

week2date("2014-W20-1", week_start = "Monday") # And back from a week to a date; the -1 is the first day of that week
```

# Converting dates/time zones

## Changing time zones with with_tz() and force_tz()

```{r}
Sys.timezone() # The computer's local time zone

time_now <- Sys.time()
time_now

time_london_real <- with_tz(time_now, "Europe/London") # Same instant, clock time changes
time_london_local <- force_tz(time_now, "Europe/London") # Same clock time, instant changes
time_london_real
time_london_local

time_london_real - time_london_local # The gap is the offset between the local time zone and London
```

```{r}
linelist %>% 
  mutate(admission_freetown = force_tz(date_time_of_admission, "Africa/Freetown"), # The data are from Sierra Leone, so declare that zone without changing the clock time
         admission_new_york = with_tz(admission_freetown, "America/New_York")) %>% # The same instants shown in another zone
  select(date_time_of_admission, admission_freetown, admission_new_york) %>% 
  head()
```

# Lagging and leading calculations

## Comparing each week to its neighbors with lag() and lead()

```{r}
weekly_counts %<>%
  mutate(cases_prev_wk = lag(n, n = 1), # The previous row's count; the first row has no previous week so it gets NA
         case_diff = n - cases_prev_wk, # Change since the previous week
         cases_next_wk = lead(n, n = 1)) # The next row's count; make sure the rows are sorted by date first because lag() and lead() only look at row position

head(weekly_counts, 10)
```

# Resources

```{r}
# lubridate: https://lubridate.tidyverse.org/
# lubridate cheatsheet: https://github.com/rstudio/cheatsheets/blob/main/lubridate.pdf
# R for Data Science, dates and times: https://r4ds.hadley.nz/datetimes.html
```
