---
title: "Ch12 pivot demo, live examples"
output:
  html_document:
    toc: yes
    df_print: kable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
# Made-up data for this demo, don't reuse it thinking it's real.
# Picks up where the demo file ends. Same data, one change per step.

pacman::p_load(rio, tidyverse)

linelist <- import("K:/R Study Group/Trenton/Ch12 Pivot/pivot_demo_data.xlsx")
```

# Same code, county instead of LHD

```{r}
# Only the first argument of count() changed. 13 rows instead of 8.
linelist %>%
  count(county, sex) %>%
  pivot_wider(names_from = sex, values_from = n, values_fill = 0)
```

# Flip which column makes the headers

```{r}
# names_from = lhd this time. 2 rows and 8 columns, and it already wraps the screen.
linelist %>%
  count(lhd, sex) %>%
  pivot_wider(names_from = lhd, values_from = n, values_fill = 0)
```

# Age group by LHD

```{r}
# Swap sex for age_group. South Heartland has zeros in two cells.
linelist %>%
  count(lhd, age_group) %>%
  pivot_wider(names_from = age_group, values_from = n, values_fill = 0)
```

# Two columns in the headers

```{r}
# names_from takes more than one column. Headers come out like Female_18-64.
linelist %>%
  count(lhd, sex, age_group) %>%
  pivot_wider(names_from = c(sex, age_group), values_from = n, values_fill = 0)
```

# Split those headers back apart

```{r}
# names_sep = "_" cuts each header at the underscore into two columns.
# 48 rows now, because the zeros became rows too.
linelist %>%
  count(lhd, sex, age_group) %>%
  pivot_wider(names_from = c(sex, age_group), values_from = n, values_fill = 0) %>%
  pivot_longer(cols = -lhd, names_to = c("sex", "age_group"), names_sep = "_", values_to = "n")
```

# id_cols decides the rows

```{r}
# id_cols says which column identifies a row. lhd is not in it, so lhd gets dropped.
linelist %>%
  count(lhd, county, sex) %>%
  pivot_wider(id_cols = county, names_from = sex, values_from = n, values_fill = 0)
```

# A prefix on the headers, then back with starts_with

```{r}
# names_prefix puts text in front of every new header.
wide <- linelist %>%
  count(lhd, sex) %>%
  pivot_wider(names_from = sex, values_from = n, values_fill = 0, names_prefix = "n_")

wide
```

```{r}
# cols = can be a helper instead of -lhd. Same 16 rows either way.
wide %>%
  pivot_longer(cols = starts_with("n_"), names_to = "sex", values_to = "n")
```

# values_from that is not a count

```{r}
# Mean age by LHD and sex, then widen. values_from = mean_age instead of n.
# NA for South Heartland men is honest. No men, no mean. So no values_fill here.
linelist %>%
  group_by(lhd, sex) %>%
  summarise(mean_age = round(mean(age))) %>%
  pivot_wider(names_from = sex, values_from = mean_age)
```

# Two values_from at once

```{r}
# values_from = c(n, mean_age) gives four columns, one per combination.
linelist %>%
  group_by(lhd, sex) %>%
  summarise(n = n(), mean_age = round(mean(age))) %>%
  pivot_wider(names_from = sex, values_from = c(n, mean_age))
```

# Percent of each LHD by sex

```{r}
# group_by(lhd) so sum(n) is that LHD's total. id_cols drops the raw n.
linelist %>%
  count(lhd, sex) %>%
  group_by(lhd) %>%
  mutate(pct = round(100 * n / sum(n))) %>%
  pivot_wider(id_cols = lhd, names_from = sex, values_from = pct, values_fill = 0)
```

# Plot straight from the long count

```{r}
# No pivot needed for the plot. The count table is already long.
linelist %>%
  count(lhd, age_group) %>%
  ggplot(aes(x = n, y = lhd, fill = age_group)) +
  geom_col()
```

# Totals row and column

```{r}
# adorn_totals is from janitor. The textbook uses it the same way.
pacman::p_load(janitor)

linelist %>%
  count(lhd, sex) %>%
  pivot_wider(names_from = sex, values_from = n, values_fill = 0) %>%
  adorn_totals(c("row", "col"))
```

# Where age_group came from

```{r}
# Rebuilding it from age with cut() and checking it matches. 11, 29, 10.
linelist %>%
  mutate(age_group_check = cut(age, breaks = c(0, 17, 64, 120), labels = c("0-17", "18-64", "65+"))) %>%
  count(age_group, age_group_check)
```
